Probabilistic Inference of Basic Stellar Parameters:\\ Application to Flickering Stars \altaffilmark{\titledag}

### 2015-10-14 version1.0/HBM.tex
small adjustments to the intro, discussion, abstract and method

We write the two models describing the relationship between $F_8$, \logg\ and
\rhostar\ as
\begin{equation}
	\log_{10}(\rho_\star) \sim \mathcal{N}(\mu = \alpha + \beta
	\log_{10}(F_8), \sigma = \sigma_{\rho}),
\end{equation}
\label{eq:rho}
and
\begin{equation}
	\log_{10}(g) \sim \mathcal{N}(\mu = \gamma + \delta \log_{10}(F_8),
	\sigma = \sigma_g).
\end{equation}
\label{eq:logg}
The free parameters of the two models are $\alpha$, $\beta$, $\sigma_{\rho}$,
$\gamma$, $\delta$, and $\sigma_g$.
These relationships can be described as Gaussian distributions with means
given by the equation of a straight line, and standard deviations which
describe the intrinsic scatter about the mean.

In order to account for the two-dimensional observational uncertainties in
this data set, we marginalize over the latent, `true' values of $F_8$,
\rhostar\ and \logg\ using the importance sampling method of
\citep{hogg:2010}\footnote{Note that while the pseudo-marginal MCMC method of
\citep{andrieu:2009} provides a truly unbiased estimate of the marginalized
likelhood, we expect any bias introduced by the importance sampling method to
be negligable.}.
In what follows we give an overview of this procedure, but encourage the reader
to refer back  to this original reference for a more detailed description of
the mathematics.
The observed values of the variables; $F_8$, \rhostar\ and \logg\ can be thought
of as a single `draw' from an underlying (assumed Gaussian) probability
distribution, with mean equal to the `true' value\footnote{We use `true' to
mean the value that would have been observed, given infinitely high
signal-to-noise.} and standard deviation equal to the observational
uncertainty.
We compute the likelihood of the data given the model, marginalized over the
true values, using importance sampling.
We sample from the posterior PDFs of the true values of the variables,
conditioned on the observed values, and compute the likelihood of each of those
samples.
% These posterior PDFs could be produced by the astronomers who provide the
% catalog.
% If posterior samples are made available, they can be used in this stage of
% the inference.
% In the event that they are not provided however, they can be generated.
% For example, the $log(g)$ values used here are point estimates of the
% posterior PDFs of $log(g)$, produced by \citet{chaplin:2014} using
% {\it Kepler} light curves and stellar models.
We would use the posterior samples generated in previous model fitting process
if they were available, however generating new samples is an acceptable
approximation provided the posteriors are Gaussian (or log-Normal) and the
priors used by the previous fitters were uninformative.
Such an approximation {\it will not be necessary} for those studies using
\rhostar\ or \logg\ calculated using the newly-calibrated relations
presented here, since we have published our posterior samples.
After producing new posterior PDF samples we then add up the individual
log-likelihoods of each sample to compute the total marginalized likelihood
for a star.
The marginalized likelihood of the whole dataset is then computed as the sum
of the log-likelihoods of each star.
The importance sampling method allows us to approximate the integral over the
latent variables.
We used flat priors for all except the $\sigma$ parameters, for which we used
priors which were flat in log-space.

Figures \ref{fig:rhostar} and \ref{fig:logg} show the data with the best-fit
models.
The shaded regions show the 1 and 2$\sigma$ confidence interval.
These intervals are representative of the uncertainty in the flicker relations,
but also of the intrinsic scatter in those relations.
% Given a measurement of flicker, the predictive probability distribution of
% \logg and \rhostar will be a Gaussian distribution

We also used an analytic form of the marginal likelihood, which accounts for
2-d uncertainties but does not allow for the modelling of the intrinsic scatter
as a function of $x$ or $y$.
For the relation between flicker and \rhostar This likelihood function can be
written as
\begin{eqnarray}
	p(\rho_*|\alpha, \beta, \sigma_{\rho}) = &
	\exp \left[-\frac{1}{2} \sum_{n=1}^N\frac{[y_n-(\alpha + \beta x_n)]^2}
	{\left[\beta \sigma_x^2 + (\sigma_y + \sigma_{\rho})^2\right]}\right]
	\\ \nonumber
	& + \beta^2 \sigma_x^2 + (\sigma_y + \sigma_{\rho})^2,
\end{eqnarray}
and similarly for \logg.
The parameter values found using this simpler likelihood function are included
in table \ref{tab:results2}.

@@

\begin{table}
\caption{Highest-likelihood parameter values with their 16th and 84th
	percentile uncertainties.}
\begin{tabular}{lcc}
\hline\hline
    Parameter & highest-likelihood value \\
    \hline
    $\alpha$         &   \\
    $\beta$          &   \\
    $\sigma_{\rho}$  &   \\
    $\gamma$         &   \\
    $\delta$         &   \\
    $\sigma_{g}$     &   \\
    \hline
\end{tabular}
\end{table}

>>>>

We write the two models describing the relationship between $F_8$, \logg\ and
\rhostar\ as
\begin{equation}
	\log_{10}(\rho_\star) \sim \mathcal{N} \left(\mu = \alpha + \beta
	\log_{10}(F_8), \sigma = \sqrt{\sigma_{\rho}^2 + \gamma F_8}\right),
\end{equation}
\label{eq:rho}
and
\begin{equation}
	\log_{10}(g) \sim \mathcal{N}\left(\mu = \delta + \epsilon \log_{10}(F_8),
	\sigma = \sqrt{\sigma_g^2 + \zeta F_8}\right).
\end{equation}
\label{eq:logg}
The free parameters of the two models are $\alpha$, $\beta$, $\gamma$
$\sigma_{\rho}$, $\delta$, $\epsilon$, $\zeta$ and $\sigma_g$.
These relations are Gaussian distributions with means
given by the equation of a straight line, and standard deviations which
describe the intrinsic scatter about the mean.

In order to account for the two-dimensional observational uncertainties in
this data set, we marginalize over the latent, `true' values of $F_8$,
\rhostar\ and \logg\ using the importance sampling method of
\citep{hogg:2010}\footnote{Note that while the pseudo-marginal MCMC method of
\citep{andrieu:2009} provides a truly unbiased estimate of the marginalized
likelhood, we expect any bias introduced by the importance sampling method to
be negligable.}.
In what follows we give an overview of this procedure, but encourage the
reader to refer back  to this original reference for a more detailed
description of the mathematics.
The observed values of the variables; $F_8$, \rhostar\ and \logg\ can be
thought of as a single `draw' from an underlying (assumed Gaussian)
probability
In what follows we give an overview of the importance sampling procedure, but
encourage the reader to refer back to the original reference for a more
detailed description of the mathematics.
The observed values of the variables; $F_8$, \rhostar\ and \logg\ can be
thought of as a single `draw' from an underlying (assumed Gaussian) probability
distribution, with mean equal to the `true' value\footnote{We use `true' to
mean the value that would have been observed, given infinitely high
signal-to-noise.} and standard deviation equal to the observational
uncertainty.
We compute the likelihood of the data given the model, marginalized over the
true values, using importance sampling.
We sample from the posterior PDFs of the true values of the variables,
conditioned on the observed values, and compute the likelihood of each of those
samples.
% These posterior PDFs could be produced by the astronomers who provide the
% catalog.
% If posterior samples are made available, they can be used in this stage of
% the inference.
% In the event that they are not provided however, they can be generated.
% For example, the $log(g)$ values used here are point estimates of the
% posterior PDFs of $log(g)$, produced by \citet{chaplin:2014} using
% {\it Kepler} light curves and stellar models.
We would use the posterior samples generated in previous model fitting process
if they were available, however generating new samples is an acceptable
approximation provided the posteriors are Gaussian (or log-Normal) and the
priors used by the previous fitters were uninformative.
Such an approximation {\it will not be necessary} for those studies using
\rhostar\ or \logg\ calculated using the newly-calibrated relations
presented here, since we have published our posterior samples.
After producing new posterior PDF samples we then add up the individual
log-likelihoods of each sample to compute the total marginalized likelihood
for a star.
The marginalized likelihood of the whole dataset is then computed as the sum
of the log-likelihoods of each star.
The importance sampling method allows us to approximate the integral over the
latent variables.
We used flat priors for all except the $\sigma$ parameters, for which we used
priors which were flat in log-space.

We also performed this inference using an analytic form of the marginal
likelihood, which accounts for 2-D uncertainties but does not allow the
intrinsic scatter to be a function of the dependent or independent variables.
For the relation between flicker and \rhostar, this likelihood function can be
written as
\begin{eqnarray}
	& p(\rho_*| F_8, \alpha, \beta, \sigma_{\rho}) =  \\ \nonumber
						      & \exp \left[-\frac{1}{2}
		\sum_{n=1}^N \frac{[F_{8n}-(\alpha + \beta \rho_{*n})]^2}
	{\left[\beta \sigma_{F8, n}^2 + (\sigma_{\rho *, n}
	+ \sigma_{\rho})^2\right]}\right]
	\\ \nonumber
	& + \beta^2 \sigma_{F8, n}^2 + (\sigma_{\rho *, n} + \sigma_{\rho})^2,
\end{eqnarray}
\label{eq:likelihood}
and similarly for \logg.
The parameter values found using this simple likelihood function are included
in table \ref{tab:results2}.
% The analytic likelihood function, equation \ref{eq:likelihood} is suitable for
% cases where the intrinsic scatter is not a function of $x$ or $y$.
Figures \ref{fig:rhostar} and \ref{fig:logg} show the data with the best-fit
models.
The shaded regions show the 1 and 2$\sigma$ confidence interval which are
representative of the intrinsic scatter in the relations.
% Given a measurement of flicker, the predictive probability distribution of
% \logg and \rhostar will be a Gaussian distribution

@@

\begin{table}
\caption{Highest-likelihood parameter values with their 16th and 84th
	percentile uncertainties.}
\begin{tabular}{lcc}
\hline\hline
    Parameter & highest-likelihood value \\
    \hline
    $\alpha$         &   \\
    $\beta$          &   \\
    $\gamma$         &   \\
    $\sigma_{\rho}$  &   \\
    $\delta$         &   \\
    $\epsilon$       &   \\
    $\zeta$          &   \\
    $\sigma_{g}$     &   \\
    \hline
\end{tabular}
\end{table}

### 2015-10-14 version1.0/discussion.tex
small adjustments to the intro, discussion, abstract and method

We have recalibrated the relation between short timescale brightness
fluctuations in the {\it Kepler} light curves of stars (flicker) with both
stellar density and surface gravity.
This is a simple `fitting a line to data' exercise, however it continues the
discussion of probabilistic modelling that is an active topic within the
fields of exoplanet and stellar astronomy.
We use Hierarchical Bayesian Modelling (HBM) to constrain the intrinsic
scatter in the relationship between flicker, surface gravity and density.
We also include the effects of the non-negligable two-dimensional observational
uncertainties by marginalizing over latent variables using importance sampling.
Although none of these methods, nor the data used here are new taken alone,
we use the example of flicker to demonstrate the ideas and the importance of
probabilistic, hierarchical inference.
Relationships between astronomical parameters are almost always
non-deterministic; an element of stochasticity effects the physical parameters
of stars so one can never perfectly predict $y$ given an observation of $x$.
We advocate a probabilistic approach in both the `fitting the model to data'
step, {\it and} when using an empirically calibrated model to predict
parameter values.
The fitting stage benefits because if the relationships between parameters are
falsely assumed to be deterministic, they will be skewed by data points with
unrepresentative uncertainties.
The prediction stage benefits from the stochastic treatment both because a
probability distribution is in many ways more representative of an observation
than a point estimate, and because posterior PDF samples can be used in
subsequent studies (provided the prior used during the fitting process is
described).

>>>>

We have recalibrated the relation between short timescale brightness
fluctuations in the {\it Kepler} light curves of stars (flicker) with both
stellar density and surface gravity, whilst including parameters to describe
the intrinsic scatter in these relationships.
We included both an additive variance term {\it and} a term that allows for
some abscissa-dependent variance, presented in table \ref{tab:table}.
The additive terms, $\sigma_\rho$ and $ \sigma_g$ are both greater than zero,
suggesting that there is an additional source of scatter in the relations, not
accounted for by the observational uncertainties alone.
This is either caused by intrinsic scatter in the physical relationship
between flicker and density and \logg, produced by some physical process that
is not accounted for in the model, or by an underestimation of the
observational uncertainties.
Interestingly, the abscissa-dependent variances, $\gamma$ and $\zeta$ are both
less than zero.
This indicates that the intrinsic scatter either decreases with increasing
flicker and decreasing stellar density and surface gravity, or that the
uncertainty underestimation is more severe for the giant stars.
Regardless of the cause of the additional `jitter' needed to describe the
relations between these parameters, the conclusions of this study are the same.
Firstly, naively modelling stochastic relationships deterministically will
lead to an underprediction of the dependent variable.
Secondly, we find no evidence for a weakening dependence at higher surface
gravities and densities, i.e. the relations between flicker, surface gravity
and stellar density are equally valid for dwarfs and giants alike.

This is a simple `fitting a line to data' exercise, however it continues the
discussion of probabilistic modelling that is an active topic within the
fields of exoplanet and stellar astronomy.
We use Hierarchical Bayesian Modelling (HBM) to constrain the intrinsic
scatter in the relationship between flicker, surface gravity and density.
We also include the effects of the non-negligable two-dimensional observational
uncertainties by marginalizing over latent variables using importance sampling.
% Although none of these methods, nor the data used here are new taken alone,
% we use the example of flicker to demonstrate the ideas and the importance of
% probabilistic, hierarchical inference.
Relationships between astronomical parameters are almost always
non-deterministic; an element of stochasticity effects the physical parameters
of stars so one can never perfectly predict $y$ given an observation of $x$.
We advocate a probabilistic approach in both the `fitting the model to data'
step, {\it and} when using an empirically calibrated model to predict
parameter values.
The fitting stage benefits because if the relationships between parameters are
falsely assumed to be deterministic, they will be skewed by data points with
unrepresentative uncertainties.
The prediction stage benefits from the stochastic treatment both because a
probability distribution is in many ways more representative of an observation
than a point estimate, and because posterior PDF samples can be used in
subsequent studies (provided the prior used during the fitting process is
described).

### 2015-10-14 version1.0/introduction.tex
small adjustments to the intro, discussion, abstract and method

Basic stellar parameters, such as effective temperature and surface gravity, 
can be inferred using one (or more) of several types of observations, such as
spectroscopy, photometry, interferometry, etc. This inference can be done by
invoking theoretical models or by building an empirical calibration library.
For example, an observed stellar spectrum could be matched against a library of 
theoretical spectra generated using stellar atmosphere models, or, against a 
library of observed spectra of ``standard stars'', serving as calibrators.
%Fundamental stellar parameters, such as mass and radius, which cannot be 
%directly inferred from the available observations can often be inferred by 
%invoking theoretical stellar models, e.g. fitting model isochrones to 
%spectroscopic measurements.
Regardless of the approach, be it theoretical or empirical, the methods used
for the inference of stellar parameters are traditionally ``deterministic''.
In this context, a deterministic model can be loosely described as one where
a particular observational input always returns a single-valued output for a 
parameter of interest. In this framework, nature itself has no variance and
our underlying model is considered to be a perfect description of reality.

An alternative approach for inferring model parameters is the 
``probabilistic'' one. Unlike the deterministic case, a single observational 
input is intepretted to be caused by a range of possible model parameters
described by a probability distribution. This statement is true even in the case
of a perfect observation of infinite signal-to-noise, since our underlying 
model itself comes with uncertainty. In recent years, there has been a shift 
towards such methods in several areas of astronomy, particularly within the 
exoplanet community. For example, \citet{wolfgang:2015} considered that
the mass-radius relationship of exoplanets is probabilistic, since a particular
sized planet could be due to a range of planet masses, cdue to unmodeled 
variances in compositions, environment and other complications. These recent 
demonstrations in exoplanetary science have prompted us to consider the need for 
treating the parent stars in the same probabilistic framework, with potential 
applications spanning many fields of astronomy.

\textbf{
The demand for probilistic stellar parameters is not only motivated by the fact
that probability distributions are far more representative of our `beliefs'
about astrophysical parameters, but also... (ruth will continue this sentence)
}

@@

\citet{cranmer:2014} demonstrated that models of stellar surface granulation 
indeed reproduce a flicker effect in close agreement with that observed by
\citet{bastien:2013}, providing a physically-plausible explanation. Since
surface gravity is highly correlated to mean stellar density (\rhostar) on
evolutionary tracks, \citet{kipping:2014} showed that flicker can be also
be used to infer \rhostar, which is more useful for exoplanet transit
analysis \citep{seager:2003}.

Whether one calibrates flicker to \logg\ or \rhostar, there are several aspects
of the problem which are attractive for our purposes of a simple demonstration
of probabilistic inference of stellar parameters. Firstly, in log-log space the
relationship is very simple, appearing to be linear \citep{kipping:2014}. 
Secondly, there is a sufficiently large number of points in the sample (439 
stars) to constrain a population-based model. Thirdly, there is significant 
excess scatter around the best-fitting relation implying that a deterministic 
model is inadequate. This is not surprising given that granulation is a complex 
and messy process for which one should not expect any parametric model to 
provide a perfect description. For these reasons, we identify the calibration of 
flicker to \logg\ and \rhostar\ as a well-posed problem to first demonstrate 
probabilistic inference in the arena of stellar characterization.

>>>>

Basic stellar parameters, such as effective temperature and surface gravity,
can be inferred using one (or more) of several types of observations, such as
spectroscopy, photometry, interferometry, etc. This inference can be performed
by invoking theoretical models or by building an empirical calibration
library.
For example, an observed stellar spectrum could be matched against a library of
theoretical spectra generated using stellar atmosphere models, or, against a
library of observed spectra of ``standard stars'', serving as calibrators.
%Fundamental stellar parameters, such as mass and radius, which cannot be
%directly inferred from the available observations can often be inferred by
%invoking theoretical stellar models, e.g. fitting model isochrones to
%spectroscopic measurements.
Regardless of the approach, be it theoretical or empirical, the methods used
for the inference of stellar parameters are traditionally ``deterministic''.
In this context, a deterministic model can be loosely described as one where
a particular observational input always returns a single-valued output for a
parameter of interest, i.e. nature itself has no variance and the underlying
model is considered to be a perfect description of reality.

An alternative approach for inferring model parameters is to allow
relationships between observables to be stochastic.
Unlike the deterministic case, a single observational input is intepreted to
be caused by a range of possible model parameters, described by a probability
distribution.
This statement is true even in the case of a perfect observation of infinite
signal-to-noise, since the underlying model itself comes with uncertainty.
In recent years, there has been a shift towards such methods in several areas
of astronomy, particularly within the exoplanet community.
For example, \citet{wolfgang:2015} considered that the mass-radius
relationship of exoplanets is stochastic, since a particular sized planet
could be have a range of planet masses due to unmodeled variances in
compositions, environment and other complications.
These recent demonstrations in exoplanetary science have prompted us to
consider the need for treating the parent stars in the same probabilistic
framework, with potential applications spanning many fields of astronomy.

The demand for probilistic stellar parameters is not only motivated by the fact
that probability distributions are far more representative of our `beliefs'
about astrophysical parameters, it also has a practical purpose.
When using data published in the astronomical literature to, for example, infer
relationships between observed parameters, that inference can be performed as
the final stage in a hierarchical treatment
\citep[see, e.g.][]{forman-mackey:2014}.
This is useful if one wants to account for multi-dimensional uncertainties by
marginalizing over the `true' parameter values upon which the noisy
observations are conditioned.
This point is discussed further in \textsection\ref{sec:HBM}.

@@

\citet{cranmer:2014} demonstrated that models of stellar surface granulation
indeed reproduce a flicker effect in close agreement with that observed by
\citet{bastien:2013}, providing a physically-plausible explanation.
Since surface gravity is highly correlated with mean stellar density (\rhostar)
on evolutionary tracks, \citet{kipping:2014} showed that flicker can be also
be used to infer \rhostar, which is more useful for exoplanet transit analysis
\citep{seager:2003}.

Whether one calibrates flicker to \logg\ or \rhostar, there are several aspects
of the problem which are attractive for our purposes of a simple demonstration
of probabilistic inference of stellar parameters.
Firstly, in log-log space the relationship is very simple, appearing to be
linear \citep{kipping:2014}.
Secondly, there is a sufficiently large number of points in the sample (439
stars) to constrain a population-based model.
Thirdly, there is significant excess scatter around the best-fitting relation
implying that a deterministic model is inadequate.
This is not surprising given that granulation is a complex and messy process
for which one should not expect any parametric model to provide a perfect
description.
Fourthly, the physical processes that produce surface granulation, of which
flicker is an observational tracer, may be more `noisy' for different types of
stars.
We will test whether flicker has greater predictive power in certain regions
of parameter space; i.e. is flicker significantly more informative for
subgiants than for dwarfs?
For these reasons, we identify the calibration of
flicker to \logg\ and \rhostar\ as a well-posed problem to first demonstrate
probabilistic inference in the arena of stellar characterization.

### 2015-10-14 version1.0/manuscript.tex
small adjustments to the intro, discussion, abstract and method

% aims
The relations between observable stellar parameters are usually assumed to be
deterministic.
That is, given an infinitely precise measurement of independant
variable, `$x$', and some model, the value of dependant variable, `y' can be
known exactly.
In practise this assumption is rarely valid and intrinsic stochasticity means
that two stars with exactly the same `$x$', will have slightly different
`$y$'s.
% methods
The relation between short-timescale brightness fluctuations (flicker) of stars
and both surface gravity \citep{bastien:2013} and stellar density
\citep{kipping:2014} are two such stochastic relations that have, until now,
been treated as deterministic ones.
We recalibrate these relations in a probabilistic framework, using
Hierarchical Bayesian Modelling (HBM) and constrain the instrinsic scatter in
the relations.
Our treatment also fully accounts for the observational uncertainties on all
variables.
% results
These new relations will provide probability distributions over values of
\logg\ or \rhostar\ for given measurements of flicker.

>>>>

% aims
The relations between observable stellar parameters are usually assumed to be
deterministic.
That is, given an infinitely precise measurement of independant
variable, `$x$', and some model, the value of dependant variable, `y' can be
known exactly.
In practise this assumption is rarely valid and intrinsic stochasticity means
that two stars with exactly the same `$x$', will have slightly different
`$y$'s.
% methods
The relation between short-timescale brightness fluctuations (flicker) of stars
and both surface gravity \citep{bastien:2013} and stellar density
\citep{kipping:2014} are two such stochastic relations that have, until now,
been treated as deterministic ones.
We recalibrate these relations in a probabilistic framework, using
Hierarchical Bayesian Modelling (HBM) and constrain the instrinsic scatter in
the relations.
Our treatment also fully accounts for the observational uncertainties on all
variables.
% results
We find evidence for additional scatter in the relationships, that cannot be
accounted for by the observational uncertainties alone.
This intrinsic scatter increases with decreasing flicker, i.e. there is more
additional scatter for subgiants than for dwarfs, suggesting that the flicker
relationships are equally (if not more) valid for dwarf stars.
